Numerical trajectory patterns of floating macroalgae

### 2021-02-16 thesis_chapter_3.Rmd
CR comments added. Updated map and morph fig added. more to come...

---
title: 'Numerical trajectory patterns of floating macroalgae'
author:
- R Coppin^[University of the Western Cape]
- C Rautenbach^[South African Weather Service]
- AJ Smit^[University of the Western Cape]
output:
  html_document:
  word_document:
    reference_docx: markdown/frontiers_template.docx
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
bibliography: references.bib
csl: markdown/frontiers.csl.txt
---

@@

#library(computeMSD)
library(ggplot2)
library(rgdal)
library(tidyverse)
library(gridExtra)
library(ggpubr)
library(ggord)
library(mapproj)
library(vegan)
library(dplyr)
library(ggpubr)
library(maps)
library(flextable)
library(officer)
library(tidync)
library(tidyverse)
library(hexbin)
library(lubridate)

@@

There is a range of objects, both natural and anthropogenic, floating in the ocean of which macroalgae are regarded as one of the most important passive dispersal mechanisms of marine taxa. Floating kelp acts as a passive dispersal mechanism for a range of marine taxa and is sometimes referred to as the 'tumble-weed' of the ocean [@edgar1987; @bushing1994; @norton1992; @helmuth1994; @holmquist1994; @smith2002; @McCormick2008]. Some macroalgae species are negatively buoyant and sink to the seafloor when detached from the substrate, while other species of macroalgae have air-filled pneumatocysts or stipes which allow the plants to reach the surface where light is more abundant. In turn, the positively buoyant pneumatocysts cause plants to float to the surface when dislodged from the substratum. The structure and number of pneumatocysts varies between species. For example species from the genera *Macrocystis*, *Sargassum*, *Ascophyllum*, and *Fucus* have thalli with many small pneumatocysts, while other kelp species such as *Nereocystis luetkeana*, *Pelagophycus porra*, *Ecklonia radiata*, *Ecklonia maxima* have a single, large pneumatocysts [@dayton1985; @smith2002; @thiel2005; @graiff2016; @batista2018]. In some cases the stipe itself is air-filled, such as with *E. maxima* and *E. radiata*. Although there have been reports of floating Chlorophyta species and some Rhodophyta species, Phaeophyceae species are the most commonly reported forms of floating algae. This is most likely because the green and red species reported floating are not actually positively buoyant, but instead are kept at the surface by gas trapped inbetween or in the thalli [@dromgoole1982; @back2000]. The giant kelp *Macrocystis pyrifera* [@helmuth1994; @kingsford1995; @hobday2000; @macaya2005; @graiff2016; @batista2018] and the bull kelp *Durvillaea antarctica* [@collins2010; @tala2013 ; @tala2017; @saunders2014; @smith2002; @wichmann2012; @batista2018] have been the focus of much of the research regarding spatial and temporal dispersal patterns, the ecological role of rafting, marine connectivity and raft-time.

Past research points to macroalgae trajectory being largely determined by prevailing wind conditions and surface currents [@thiel2005, @thiel2005; @hobday2000; @fraser2011; @rothausler2011; @rothausler2011]. Although ocean currents are regarded as the primary influence, the relative importance of wind versus surface current is still not known; although the role of wind has been recognised as important in several studies. For example, a study by @harrold1989 investigated the seasonal trajectories of radio-transmitter tagged *M. pyrifera* in nearshore Monterrey Bay. The results showed that kelp rafts with little surface area exposed to the wind were largely driven by a combination of wind and wind waves, however the relative importance of wind and wind waves was not clear. In addition, the tagged kelp trajectories were more consistent with the formation of eddies during winter. Previous studies have identified wind as an important mechanism of dispersal in wind dominated ocean systems. For example the subAntartic latitudes the West Wind Drift causes continuous unidirectional surface flow and is regarded as an important potential mechanism for dispersal of floating kelp. Other studies have used genetic approaches to determine macroalgae raft trajectory characteristics by inferring source location from genetically similar populations [@nikula2013]. For example, a study by @fraser2011 on the rafting capabilities of *Durvillaea antarctica* used a combination of population genetics and relative age estimate of ‘goosebarnacles’ attached to the raft. The presence of goosebarnacles suggests a long raft time as these species have a slow growth rate; while the genetic analyses showed these species are able to raft up to $\sim$ 390km from their local origin. The authors suggested that wind and water-movement were the primary influences of trajectory, however, this was only inferred from the genetic results and local climatology data [@fraser2011].

@@

To accurately determine the trajectory of marine macroalgae these aspects need to be taken into account. Past research by @allen1999 and @breivik2011 have provided estimates of drag for various objects based on experimental work, which consisted of both direct and indirect methods of the Leeway model. However, this work does not consider biological material such as macroalgae. Although drag estimates do exist, these have been applied for macroalgae not detached from the substratum and are regarded as fixed-point estimates. Although various approaches have been used in the past to investigate floating macroalagae trajectory, very few studies have employed the use of Langrangian trajectory modeling. Furthermore, non of the existing studies using this modeling approach have considered macroalgal morphology cross-sectional area (i.e. shape) as an aspect of drag and ultimately trajectory. Studies by @brooks2019, @putman2018, @putman2020 used lagrangian approaches to investigate the effects of inertia, raft-size, and windage on macroalgae trajectory. @brooks2019 used a custom growth model to estimate changes in biomass and ultimately radial size, while a customised Hybrid Coordinate Ocean Model (HYCOM) for the trajectory simulations was used. The results showed that trajectory of pelagic *Sargussum* was significantly influenced by inertia and the radial size of the rafts. Other work by @putman2020 investigated the effect of including windage in macroalgae trajectory simulations. The authors showed that including ad-hoc windage factors greatly improved accuracy regarding entrainment or expulsion from eddies when compared to tracked sargassum mats in the ocean. Langrangian trajectory modeling is a useful tool, however, the various models available often assume the object to be a spherical particle, which is obviously not the case for both biological and anthropogenic forms of marine debris in nature. The radial size included by @brooks2019 is an improvement, however this approach does not take into account the complex morphology on an individual level into account.

@@

# Methods

## Study site

The region of interest for this study was the Cape Peninsula in the Western Cape (Figure 1A). A release site was chosen for this study, Kommetjie, which is located on the Atlantic side of the Cape Peninsula (Figure 2B). The hydrodynamics in the study region are driven by a complex interaction of wind and waves. which are also influenced by larger oceanic processes In South Africa, the larger processes can largely be attributed to the Agulhas Current (AC) and the Benguela Current (BC). The AC is a component of the South-west Indian Ocean sub-gyre, flowing predominantly south-westward following the continental shelf edge and eventually retroflecting eastward back into the South Indian Ocean. The region where the retroflection occurs is known as the Agulhas retroflection and consists of some of the highest levels of mesoscale variability [@garzoli1996]. The variability results in shedding of rings and filaments known as the Agulhas Rings and have an average diameter of ~324km [@lutjeharms1988; @lutjeharms2006; @@lutjeharms2007]. The Agulhas Rings are the main contributors to the Agulhas leakage which is the transport of warm water from the Indian Ocean to the South Atlantic Ocean [@garzoli1996; @lutjeharms2006; @beal2015]. The Agulhas leakage is also a source of meso-scale variability within the Benguela region as anticyclonic rings are shed and drift northwestward along the southern Benguela slope region [@rubio2009]. The Benguela region is characterised by strong coastal upwelling cells which a primarily driven by equatorward winds [@shannon1996; @hardman2003], and is known as the Benguela upwelling region. The Benguela upwelling region starts at 27$^\circ$S and extends southward to 35$^\circ$S [@shannon1996]. The region is also characterised by high amounts of mesoscale variability in the form of eddies and filaments[@blanke2002; @blanke2005; @rubio2009]. The Benguela inshore region is dominated by a northwestward flow as a result of topographical steering, wind stress and interactions with passing Agulhas rings and eddies [@veitch2010].

@@

map_chp3A <- rasterGrob(as.raster(readPNG("map/chp3_mapA.png")))
map_chp3B <- rasterGrob(as.raster(readPNG("map/chp3_mapB.png")))

ggarrange(map_chp3A, map_chp3B, ncol = 1, nrow = 2, labels = "AUTO")
```

@@

$$X(t+\Delta t) = X(t) + \int_{t}^{t+\Delta t} v(x, \tau)d \tau + \Delta X_b(t)$$
where $X$ is the three-dimensional position of a particle, $v(x,t)$ is the three-dimensional velocity field at the location in the ocean model, $X_b(t)$ is a change in position due to 'behavior' of the particle. This can range from swimming in fish to sinking or beaching. IN order to investigate the effects of drag components of trajectory of floating macroalgae, this study incorporated drag as the custom behavior.

This study simulated two scenarios; one where the plant is fully submerged and only exposed to hydrodynamic drag; and the other where parts of the plants are partially exposed resulting in a combination of hydrodynamic and air form drag. Drag based on shape based coefficients were used in the calculations for determining the overall kelp velocity vectors. A momentum energy equation was used to calculate the drag forces for the relevant simulation. This approach has been employed in modeling iceberg drift trajectory [@andersson2017; @lichey2001; @eik2009] and was adapted to suit this particular study. The energy momentum equation used to calculate the drag force exerted on the virtual kelp particle was, $$m({d_u/d_t})=F_d + F_a + F_f$$
where $m$ is kelp mass, $d_u$ the virtual kelp particle velocity, $F_d$ the hydrodynamic drag, $F_a$ is the wind drag, and $F_f$ the surface-current flow. The assumption made was that the mass of the kelp did not change for each virtual kelp particle over the course of the simulations. Hydrodynamic drag, wind drag and surface current-flow are two-dimensional vector quantities and were calculated for each time step for each simulation.

## Hydrodynamic and wind model

Both passive and kelp particles were simulated around the South African coastline within hind-cast outputs from the Copernicus Marine Environment Monitoring Service. The Copernicus outputs used in this study contains 3D daily current information from the top layer to the bottom (Global Analysis Forecast PHY_001_024). The model has a spatial resolution of 0.08$^\circ$ and is interpolated on a Arakawa C native grid.

@@

Virtual particles were released with no drag or wind effects from the study site in order to act as a basis for comparison to simulations with varying degrees of hydrodynamic and wind drag, and are referred to hereafter as 'passive particles'. All simulations have the same release location and were run for 30 days with hourly outputs using a fourth order Runge-Kutte advection scheme. Brownian motion was also incorporated in order to account for any stochasticity not resolved by the model. These are standard practices for Lagrangian trajectory studies.

@@

In order to determine the influence of cross-sectional area on drag components, simulations were performed for the minimum, mean and maximum cross-sectional areas for Kommetjie kelp forest population (Figure 1). In order to determine the direct effect of wind on trajectory, simulations were run for varying magnitudes of wind exposure for each type of cross-sectional area. The virtual particles released with the hydrodynamic or wind drag components are referred to as 'kelp particles'.

## Model inputs
### Cross-sectional area

In order to incorporate hydrodynamic and wind drag, the cross-sectional area of the kelp was calculated first. Known geometric shapes reflecting the relevant plant sections were used to estimate the surface area for various parts of the plant, for details please refer to table 2. The dimensional data needed was estimated in cases were data was not available for that particular morphological characteristic. The bulb/pneumatocyst is a highly variable morphological characteristic and in some cases can appear absent, the same is true for the holdfast. Therefore, the cross-sectional area of the bulb/pneumatocyst was not considered in the calculation of overall cross-sectional area. In terms of the holdfast, a standard cross-sectional area was used for all simulations. The cross-sectional area calculated was site specific and the dimensions needed were garnered from morphology data from a previous study [@coppin2020]. The minimum, mean and maximum were calculated for each morphological characteristic (see appendix), which were used for calculating the overall minimum, mean and maximum cross-sectional areas needed to run the various simulations. The trajectory of these "types" of cross-sectional areas (minimum, mean and maximum overall cross-sectional area) were used to determine the influence of drag, both hydrodynamic and wind, on overall trajectory.

@@

area_table <- read.csv("tables/chapter_3/area_table.csv", sep = ";")
area_table <- flextable(data = area_table)
area_table <- theme_zebra(area_table, odd_header = "transparent", even_header = "transparent")
area_table <- align(x= area_table, align = "center", part = "all")
area_table <- set_header_labels(area_table, Plant.characteristic = "Plant characteristic", Approximate.shape = "Approximate shape", Equation.used = "Equation Used", Plant.dimensions = "Plant dimensions")
area_table <- set_header_labels(area_table, Ac.Minimum = "Ac Minimum", Ac.Mean = "Ac Mean", Ac.Maximum = "Ac Maximum")
area_table <- set_table_properties(area_table, width = 1, layout = "autofit")
area_table <- set_caption(area_table, "Table 2:Summary of the atrributes and estimates used to calculate the overall surface area of Kommetjie kelp individuals used in the various simulation.")
area_table
```

The drag coefficient for a spherical particle was used in the calculation of hydrodynamic drag as the trajectory model assumes the particles are spherical. The drag equation used was: $F_d = 0.5 \rho A_c C_d U^2$, where $\rho$ is the density of seawater or air, $A_c$ is the cross-sectional area exposed to hydrodynamic or wind drag, $C_d$ is the drag coefficient for a spherical particle and $U$ is the surface velocity vector of the flow-/wind- field. Hydrodynamic and wind drag components were calculated separately for both the meridonal and zonal velocity vectors. Since drag force is dependent on the velocity vectors which vary with time, the meridonal and zonal velocities were interpolated and used in the drag force calculation for each time-step. The same approach was used for the wind drag force. The equation that was used to calculate hydrodynamic drag, was also used to calculate wind. To test the different magnitudes that hydrodynamic and wind drag might play in the overall trajectory, simulations were run with varying degrees of drag exposure scenarios. These scenarios were meant to act as proxy for buoyancy, which determines the surface area exposed to hydrodynamic and wind drag. These estimates were expressed as percentages which were used to calculate hydrodynamic and wind drag for applicable simulation. For example, if 80% of the plant is submerged and 20% of above water, then the overall hydrodynamic and wind drag forces would be 85% and 15% of the total drag forces respectively. The exposure scenarios used in this study were 100%, 95%, 90%, and 85% for hydrodynamic drag; 0 %, 5%, 10%, and 15% for wind (see table 1).

## Analysis

@@

wwdrag_plot <- ggplot()+
  geom_path(data = p_RK4_mean, aes(x = lon, y = lat), color = 'black')+
  geom_path(data = Min_Kfloat_H095_W05_mean, aes(x = lon, y = lat), color = 'snow4')+
  geom_path(data = Min_Kfloat_H090_W10_mean, aes(x = lon, y = lat), color = 'green')+
  geom_path(data = Min_Kfloat_H085_W15_mean, aes(x = lon, y = lat), color = 'pink')+
  geom_path(data = Mean_KFloat_H095_W05_mean, aes(x = lon, y = lat), color = 'yellow')+
  geom_path(data = Mean_KFloat_H090_W10_mean, aes(x = lon, y = lat), color = 'orange')+
  geom_path(data = Mean_KFloat_H085_W15_mean, aes(x = lon, y = lat), color = 'red')+
  geom_path(data = Max_Kfloat_H095_W05_mean, aes(x = lon, y = lat), color = 'blue')+
  geom_path(data = Max_Kfloat_H090_W10_mean, aes(x = lon, y = lat), color = 'orange')+
  geom_path(data = Max_Kfloat_H085_W15_mean, aes(x = lon, y = lat), color = 'red')+
  xlab("Longtitude") + 
  ylab("Latitude") + 
  ggtitle("") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_bw()

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap="Figure 2: Comparison of mean trajectories. Plot A is the comparison of mean trajectories for only hydrodynamic drag and plot B is the comparison of mean trajectories for simulations with varying degrees of hydrodynamic and wind drag."}
ggarrange(waterdrag_plot, wwdrag_plot, labels = "AUTO")
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Figure 3: Comparison of density of particles within each grid cell for the end run time of each scenario. Plots A-D repressent the minimum, plots E-H are the mean and plots I-L the maximum cross-sectional areas. The bottom plot is the density plot for the passive simulation.', fig.height=10.2, fig.width=7.5}

>>>>

---
title: 'Numerical trajectory patterns of floating macroalgae'
author:
- R Coppin^[University of the Western Cape]
- C Rautenbach^[South African Weather Service]
- AJ Smit^[University of the Western Cape]
output:
  word_document:
    reference_docx: markdown/frontiers_template.docx
  html_document:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
bibliography: references.bib
csl: markdown/frontiers.csl.txt
---

@@

#library(computeMSD)
library(ggplot2)
library(rgdal)
library(gridExtra)
library(ggpubr)
library(ggord)
library(mapproj)
library(vegan)
library(dplyr)
library(ggpubr)
library(maps)
library(flextable)
library(officer)
library(tidync)
library(tidyverse)
library(hexbin)
library(lubridate)

@@

There are a range of objects, both natural and anthropogenic, floating in the ocean of which macroalgae are regarded as one of the most important passive dispersal mechanisms of marine taxa. Floating kelp acts as a passive dispersal mechanism for a range of marine taxa and is sometimes referred to as the 'tumble-weed' of the ocean [@edgar1987; @bushing1994; @norton1992; @helmuth1994; @holmquist1994; @smith2002; @McCormick2008]. Some macroalgae species are negatively buoyant and sink to the seafloor when detached from the substrate, while other species of macroalgae have air-filled pneumatocysts or stipes which allow the plants to reach the surface where light is more abundant. In turn, the positively buoyant pneumatocysts cause plants to float to the surface when dislodged from the substratum. The structure and number of pneumatocysts varies between species. For example species from the genera *Macrocystis*, *Sargassum*, *Ascophyllum*, and *Fucus* have thalli with many small pneumatocysts, while other kelp species such as *Nereocystis luetkeana*, *Pelagophycus porra*, *Ecklonia radiata*, *Ecklonia maxima* have a single, large pneumatocysts [@dayton1985; @smith2002; @thiel2005; @graiff2016; @batista2018]. In some cases the stipe itself is air-filled, such as with *E. maxima* and *E. radiata*. Although there have been reports of floating Chlorophyta species and some Rhodophyta species, Phaeophyceae species are the most commonly reported forms of floating algae. This is most likely because the green and red species reported floating are not actually positively buoyant, but instead are kept at the surface by gas trapped inbetween or in the thalli [@dromgoole1982; @back2000]. The giant kelp *Macrocystis pyrifera* [@helmuth1994; @kingsford1995; @hobday2000; @macaya2005; @graiff2016; @batista2018] and the bull kelp *Durvillaea antarctica* [@collins2010; @tala2013 ; @tala2017; @saunders2014; @smith2002; @wichmann2012; @batista2018] have been the focus of much of the research regarding spatial and temporal dispersal patterns, the ecological role of rafting, marine connectivity and raft-time.

Past research points to macroalgae trajectory being largely determined by prevailing wind conditions and surface currents [@thiel2005; @hobday2000; @fraser2011; @rothausler2011; @rothausler2011]. Although ocean currents are regarded as the primary influence, the relative importance of wind versus surface current is still unknown although the role of wind has been recognised as important in several studies. For example, a study by @harrold1989 investigated the seasonal trajectories of radio-transmitter tagged *M. pyrifera* in nearshore Monterrey Bay. The results showed that kelp rafts, with little surface, area exposed to the wind were largely driven by a combination of wind and wind waves, however the relative importance of wind and wind waves was not clear. In addition, the tagged kelp trajectories were more consistent with the formation of eddies during winter. Previous studies have identified wind as an important mechanism of dispersal in wind dominated ocean systems. For example, the subAntartic latitudes the West Wind Drift causes continuous unidirectional surface flow and is regarded as an important potential mechanism for dispersal of floating kelp. Other studies have used genetic approaches to determine macroalgae raft trajectory characteristics by inferring source location from genetically similar populations [@nikula2013]. For example, a study by @fraser2011 on the rafting capabilities of *Durvillaea antarctica* used a combination of population genetics and relative age estimate of ‘goosebarnacles’ attached to the raft. The presence of goosebarnacles suggests a long raft time as these species have a slow growth rate; while the genetic analyses showed these species are able to raft up to $\sim$ 390km from their local origin. The authors suggested that wind and water-movement were the primary influences of trajectory, however, this was only inferred from the genetic results and local climatology data [@fraser2011].

@@

To accurately determine the trajectory of marine macroalgae these aspects need to be taken into account. Past research by @allen1999 and @breivik2011 have provided estimates of drag for various objects based on experimental work, which consisted of both direct and indirect methods of the Leeway model. However, this work does not consider biological material such as macroalgae. Although drag estimates do exist, these have been applied for macroalgae not detached from the substratum and are regarded as fixed-point estimates. Although various approaches have been used in the past to investigate floating macroalagae trajectory, very few studies have employed the use of Langrangian trajectory modeling. Furthermore, none of the existing studies using this modeling approach have considered macroalgal morphology cross-sectional area (i.e. shape) as an aspect of drag and ultimately trajectory. Studies by @brooks2019, @putman2018, @putman2020 used lagrangian approaches to investigate the effects of inertia, raft-size, and windage on macroalgae trajectory. @brooks2019 used a custom growth model to estimate changes in biomass and ultimately radial size, while a customised Hybrid Coordinate Ocean Model (HYCOM) for the trajectory simulations was used. The results showed that trajectory of pelagic *Sargussum* was significantly influenced by inertia and the radial size of the rafts. Other work by @putman2020 investigated the effect of including windage in macroalgae trajectory simulations. The authors showed that including ad-hoc windage factors greatly improved accuracy regarding entrainment or expulsion from eddies when compared to tracked sargassum mats in the ocean. Langrangian trajectory modeling is a useful tool, however, the various models available often assume the object to be a spherical particle, which is obviously not the case for both biological and anthropogenic forms of marine debris in nature. The radial size included by @brooks2019 is an improvement, however this approach does not take into account the complex morphology on an individual level into account.

@@

# Methods
## Study site

The region of interest for this study was the Cape Peninsula in the Western Cape (Figure 1A). A release site was chosen for this study, Kommetjie, which is located on the Atlantic side of the Cape Peninsula (Figure 2B). The hydrodynamics in the study region are driven by a complex interaction of wind and waves, which are also influenced by larger oceanic processes. In South Africa, the larger processes can largely be attributed to the Agulhas Current (AC) and the Benguela Current (BC)[@lutjeharms2000;@lutjeharms2007;@garzoli1996;@Hutchings2009;@rubio2009]. The AC is a component of the South-west Indian Ocean sub-gyre, flowing predominantly south-westward following the continental shelf edge and eventually retroflecting eastward back into the South Indian Ocean [@@lutjeharms2000;@lutjeharms2007]. The region where the retroflection occurs is known as the Agulhas retroflection and consists of some of the highest levels of mesoscale variability [@garzoli1996]. The variability results in shedding of rings and filaments known as the Agulhas Rings and have an average diameter of ~324km [@lutjeharms1988; @lutjeharms2006; @lutjeharms2007]. The Agulhas Rings are the main contributors to the Agulhas leakage which are the transport of warm water from the Indian Ocean to the South Atlantic Ocean [@garzoli1996; @lutjeharms2006; @beal2015]. The Agulhas leakage is also a source of meso-scale variability within the Benguela region as anticyclonic rings are shed and drift northwestward along the southern Benguela slope region [@rubio2009]. The Benguela region is characterised by strong coastal upwelling cells which are primarily driven by equatorward winds [@shannon1996; @hardman2003], and is known as the Benguela upwelling region. The Benguela upwelling region starts at 27$^\circ$S and extends southward to 35$^\circ$S [@shannon1996]. The region is also characterised by high amounts of mesoscale variability in the form of eddies and filaments [@blanke2002; @blanke2005; @rubio2009]. The Benguela inshore region is dominated by a northwestward flow as a result of topographical steering, wind stress and interactions with passing Agulhas rings and eddies [@veitch2010].

@@

#map_chp3A <- rasterGrob(as.raster(readPNG("map/chp3_mapA.png")))
#map_chp3B <- rasterGrob(as.raster(readPNG("map/chp3_mapB.png")))
#ggarrange(map_chp3A, map_chp3B, ncol = 2, nrow = 1, labels = "AUTO", align = c("h"))
map_chp3_combined <- rasterGrob(as.raster(readPNG("map/chp3_map_combined_hz.png")))
ggarrange(map_chp3_combined)
```

@@

$$X(t+\Delta t) = X(t) + \int_{t}^{t+\Delta t} v(x, \tau)d \tau + \Delta X_b(t)$$
where $X$ is the three-dimensional position of a particle, $v(x,t)$ is the three-dimensional velocity field at the location in the ocean model at particle point in time, $t$ is time, $X_b(t)$ is a change in position due to 'behavior' of the particle at a particular point in time. This can range from swimming in fish to sinking or beaching. In order to investigate the effects of drag components of trajectory of floating macroalgae, this study incorporated drag as the custom behavior.

This study simulated two scenarios; one where the plant is fully submerged and only exposed to hydrodynamic drag; and the other where parts of the plants are partially exposed resulting in a combination of hydrodynamic and air form drag. Drag based on shape-based coefficients were used in the calculations for determining the overall kelp velocity vectors. A momentum energy equation was used to calculate the drag forces for the relevant simulation. This approach has been employed in modeling iceberg drift trajectory [@andersson2017; @lichey2001; @eik2009] and was adapted to suit this particular study. The energy momentum equation used to calculate the drag force exerted on the virtual kelp particle was, $$m({d_u/d_t})=F_d + F_a + F_f$$
where $m$ is kelp mass, $d_u$ the virtual kelp particle velocity, $F_d$ the hydrodynamic drag, $F_a$ is the wind drag, $F_f$ the surface-current flow, and $d_t$ is the time step. The assumption made was that the mass of the kelp did not change for each virtual kelp particle over the course of the simulations. Hydrodynamic drag, wind drag and surface current-flow are two-dimensional vector quantities and were calculated for each time step for each simulation.

@@

Virtual particles were released with no drag or wind effects from the study site in order to act as a reference for comparison to simulations with varying degrees of hydrodynamic and wind drag, and are referred to hereafter as 'passive particles'. All simulations have the same release location and were run for 30 days with hourly outputs using a fourth order Runge-Kutte advection scheme [@North2009; @VanSebille2018]. Brownian motion was also incorporated in order to account for any stochasticity not resolved by the model. These are standard practices for Lagrangian trajectory studies [@North2009;@delandmeter2019; @VanSebille2018]. To test the different magnitudes that hydrodynamic and wind drag might play in the overall trajectory, simulations were run with varying degrees of drag exposure scenarios. These scenarios were meant to act as proxy for buoyancy, which determines the surface area exposed to hydrodynamic and wind drag. These estimates were expressed as percentages which were used to calculate hydrodynamic and wind drag for applicable simulation. For example, if 85% of the plant is submerged and 15% of above water, then the overall hydrodynamic and wind drag forces would be 85% and 15% of the total drag forces respectively. The exposure scenarios used in this study were 100%, 95%, 90%, and 85% for hydrodynamic drag; 0 %, 5%, 10%, and 15% for wind (see table 1).

@@

In order to determine the influence of cross-sectional area on drag components, simulations were performed for the minimum, mean and maximum cross-sectional areas for Kommetjie kelp forest population (Figure 1). In order to determine the direct effect of wind on trajectory, simulations were run for varying magnitudes of wind exposure for each type of cross-sectional area. The magnitudes of exposure where based on personal observations of drifting *E. maxima* from previous studies and fieldwork. The virtual particles released with the hydrodynamic or wind drag components are referred to as 'kelp particles'.

## Model inputs
### Ocean and wind model

Both passive and kelp particles were simulated around the South African coastline within hind-cast outputs from the Copernicus Marine Environment Monitoring Service. The Copernicus outputs used in this study contains 3D daily current information from the top layer to the bottom (Global Analysis Forecast PHY_001_024). The model has a spatial resolution of 0.08$^\circ$ and is interpolated on an Arakawa C native grid.

The drag coefficient for a spherical particle was used in the calculation of hydrodynamic drag as the trajectory model assumes the particles are spherical. The drag equation used was: $F_d = 0.5 \rho A_c C_d U^2$, where $\rho$ is the density of seawater or air, $A_c$ is the cross-sectional area exposed to hydrodynamic or wind drag, $C_d$ is the drag coefficient for a spherical particle and $U$ is the surface velocity vector of the flow-/wind- field. Hydrodynamic and wind drag components were calculated separately for both the meridonal and zonal velocity vectors. Since drag force is dependent on the velocity vectors which vary with time, the meridonal and zonal velocities were interpolated and used in the drag force calculation for each time-step. The same approach was used for the wind drag force. The equation that was used to calculate hydrodynamic drag, was also used to calculate wind.

### Cross-sectional area

In order to incorporate hydrodynamic and wind drag, the cross-sectional area of the kelp was calculated first. Known geometric shapes reflecting the relevant plant sections were used to estimate the surface area for various parts of the plant, for details please refer to figure 2. The dimensional data needed was estimated in cases were data was not available for that particular morphological characteristic. The bulb/pneumatocyst is a highly variable morphological characteristic and in some cases can appear absent, the same is true for the holdfast. Therefore, the cross-sectional area of the bulb/pneumatocyst was not considered in the calculation of overall cross-sectional area. In terms of the holdfast, a standard cross-sectional area was used for all simulations.

```{r message=FALSE, warning=FALSE, include=TRUE, echo=FALSE, fig.height=4, fig.width=7, fig.cap="Figure 2:Diagram of the *E.maxima* morphological characteristics considered in this study, along with the associated geometric shapes used to calculate the overall surface area of each morphological characteristic."}

kelp_shape_figure <- rasterGrob(as.raster(readPNG("figures/chapter_3/chp3_shapedrawing.png")))
ggarrange(kelp_shape_figure)

The cross-sectional area calculated was site specific and the dimensions needed were garnered from morphology data from a previous study [@coppin2020]. The minimum, mean and maximum were calculated for each morphological characteristic, which were used for calculating the overall minimum, mean and maximum cross-sectional areas needed to run the various simulations (see table 1). The trajectory of these "types" of cross-sectional areas (minimum, mean and maximum overall cross-sectional area) were used to determine the influence of drag, both hydrodynamic and wind, on overall trajectory.

@@

area_table <- read.csv("tables/chapter_3/area_table.csv", sep = ";")
area_table <- flextable(data = area_table)
area_table <- theme_zebra(area_table, odd_header = "transparent", even_header = "transparent")
area_table <- align(x= area_table, align = "center", part = "all")
area_table <- set_header_labels(area_table, Plant.characteristic = "Plant characteristic", Approximate.shape = "Approximate shape", Equation.used = "Equation Used", Plant.dimensions = "Plant dimensions")
area_table <- set_header_labels(area_table, Ac.Minimum = "Ac Minimum", Ac.Mean = "Ac Mean", Ac.Maximum = "Ac Maximum")
area_table <- set_table_properties(area_table, width = 1, layout = "autofit")
area_table <- set_caption(area_table, "Table 2:Summary of the atrributes and estimates used to calculate the overall surface area of Kommetjie kelp individuals used in the various simulation.")
area_table
```
## Analysis

@@

wwdrag_plot <- ggplot()+
  geom_path(data = p_RK4_mean, aes(x = lon, y = lat), color = 'black')+
  geom_path(data = Min_Kfloat_H095_W05_mean, aes(x = lon, y = lat), color = 'grey')+
  geom_path(data = Min_Kfloat_H090_W10_mean, aes(x = lon, y = lat), color = 'green')+
  geom_path(data = Min_Kfloat_H085_W15_mean, aes(x = lon, y = lat), color = 'pink')+
  geom_path(data = Mean_KFloat_H095_W05_mean, aes(x = lon, y = lat), color = 'yellow')+
  geom_path(data = Mean_KFloat_H090_W10_mean, aes(x = lon, y = lat), color = 'orange')+
  geom_path(data = Mean_KFloat_H085_W15_mean, aes(x = lon, y = lat), color = 'red')+
  geom_path(data = Max_Kfloat_H095_W05_mean, aes(x = lon, y = lat), color = 'blue')+
  geom_path(data = Max_Kfloat_H090_W10_mean, aes(x = lon, y = lat), color = 'orange')+
  geom_path(data = Max_Kfloat_H085_W15_mean, aes(x = lon, y = lat), color = 'red')+
  xlab("Longtitude") + 
  ylab("Latitude") + 
  ggtitle("") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_bw()

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap="Figure 2: Comparison of mean trajectories. Plot A is the comparison of mean trajectories for only hydrodynamic drag and plot B is the comparison of mean trajectories for simulations with varying degrees of hydrodynamic and wind drag. In plot A, the black line represents simulation 1, the blue line 2, the orange line 6 and the red line 10. In plot B, the black line represents simulation 1, the grey line simulation 3, the green 4, the pink 5, the yellow 7, orange 8, red 9, blue 11, bronze 12, and cherry 13."}
ggarrange(waterdrag_plot, wwdrag_plot, labels = "AUTO")
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Figure 3: Comparison of density of particles within each grid cell for the end run time of each scenario. Plots A-D represents the minimum, plots E-H are the mean and plots I-L the maximum cross-sectional areas. The bottom plot is the density plot for the passive simulation.', fig.height=10.2, fig.width=7.5}
